---
title: "Ethnic Fractionalization and Political Rights of Participation"
author: "Diego Fernandez Fernandez and Isabelle Miller"
date: "2 May 2016"
output: beamer_presentation
---

```{r setup, include=FALSE}
load("workspacefinal.RData")
library("rworldmap")
library("stargazer")
knitr::opts_chunk$set(echo = FALSE)
```

## Merkel's Concept of Embedded & Defective Democracies

Internal and external embeddedness 

- External: spheres of enabling conditions that protect from         destabilization.

- Internal: independence and interdependence of five "partial regimes". We focus on political rights of participation - freedom of speech and opinion + freedom of association, demonstration, and petition.

## Paul Collier's Research on Ethnic Fractionalization and Conflict

In developed democracies:

- higher ethnic fractionalization can play a stabilizing role 

In authoritarian regimes:

- "dominance" increases probability of violent conflict

- pattern of fractionalization without dominance can also be stabilizing

## Research Question and Variables
-Does ethnic fractionalization have an effect on freedom of speech and opinion?

-Hypothesis: the more ethnically fractionalized a country is, the higher its negative impact on political rights if income inequality is high. 

-DV - pressfreedom; IV - ethnicfrac + gini + ethnicfrac*gini + GDPpercapita + langfrac + judindep

__PoliticalRightsi= B0 + B1EthnicFraci + B2IncomeIneqi + B3EthnicFracxIncomeIneqi + B4GDPcapitai + B5Judiciaryindependencei + +B6Linguisticfraci + Errori__

## Press Freedom 2002
```{r, echo=FALSE}

#pressfreedom 2002
mapCountryData(sPDF, nameColumnToPlot='pressfreedom', mapTitle= '',
               colourPalette = c("darkgreen", "green", "yellow", "orange", "darkorange", "red"),
               oceanCol = 'lightblue',
               missingCountryCol='white',
               catMethod = 'fixedWidth',
               numCats = 6,
               borderCol='black')

```

## Press Freedom 2014
```{r, echo=FALSE}
#pressfreedom 2014
mapCountryData(sPDF, nameColumnToPlot='pressfreedom', mapTitle= '',
               colourPalette = c("darkgreen", "green", "yellow", "orange", "darkorange", "red"),
               oceanCol = 'lightblue',
               missingCountryCol='white',
               catMethod = 'fixedWidth',
               numCats = 6,
               borderCol='black')
```

## Ethnic Fractionalization 2002
```{r, echo=FALSE}
#ethnicfrac 2002 (higher = more fractionalized)
mapCountryData(sPDF, nameColumnToPlot='ethnicfrac', mapTitle= '',
               colourPalette = c("light green", "green", "dark green", "black"),
               oceanCol = 'lightblue',
               missingCountryCol='white',
               catMethod = 'fixedWidth',
               numCats = 4,
               borderCol='black')

```

##Ethnic Fractionalization 2014

```{r, echo=FALSE}
#ethnicfrac 2014
mapCountryData(sPDF, nameColumnToPlot='ethnicfrac', mapTitle= '',
               colourPalette = c("light green", "green", "dark green", "black"),
               oceanCol = 'lightblue',
               missingCountryCol='white',
               catMethod = 'fixedWidth',
               numCats = 4,
               borderCol='black')

```

## Methods / Model

-Descriptive statistics

-Cross-sectional OLS regressions for a couple of years. 
(Although we have data for the period 2002-2014, we discarded panel data because our main variables of interest do not vary so much over that time)

## Preliminary Regression Results

```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, tidy=TRUE, results="asis"}

CV2008 <- subset(CombinedVars2, year == 2008)
CV2010 <- subset(CombinedVars2, year == 2010)

M1_2008 <- lm(pressfreedom ~ ethnicfrac + Ginicoef + ethnicfrac:Ginicoef + GDPpercapita + langfrac + judindep, data = CV2008)

M2_2010 <- lm(pressfreedom ~ ethnicfrac + Ginicoef + ethnicfrac:Ginicoef + GDPpercapita + langfrac + judindep, data = CV2010)

labels_IVs <- c("Ethnic Fractionalization", "Gini coefficient", "Ethnicfrac*Gini", "GDP per capita", "Judicial Independence", "Linguistic Fractionalization", ("Intercept"))
label_DV <- c("Freedom of the press")

stargazer::stargazer(M1_2008, M2_2010, dep.var.labels=label_DV, covariate.labels = labels_IVs, title = "Regression Results: Year 2008 (1) and Year 2010 (2)", digits = 2, type = "latex", header = FALSE, font.size = "tiny")

```

## What's next?

-Finding the best way to deal with variables of interest which are nearly time-invariant

-Multiple imputation

-Including lagged press freedom as independent variable

-Interpretation of interaction term

-different groupings (defective vs. embedded, dropping outliers or non-OECD)