---
title: "Ethnic Diversity and Political Rights"
authors: "Diego Fernandez Fernandez and Isabelle Miller"
date: "13 May 2016"
output: html_document
bibliography: CitationsFinal.bib
---


````{r setup, include=FALSE}
load("workspacefinal.RData")

#call packages
library("ggmap")
library("maptools")
library("ggplot2")
library("rworldmap")
library("sp")
library("dygraphs")
library("stargazer")
library("knitr")
library("plm")

##prepare new datasets
#subset years
CV2008 <- subset(CombinedVars2, year == 2008)
CV2010 <- subset(CombinedVars2, year == 2010)

#drop rows that are NA in column 7 (ginicoef)
CV08_Gini <- CV2008[complete.cases(CV2008[,7]),]
CV10_Gini <- CV2010[complete.cases(CV2010[,7]),]

#combine 2008 and 2010 by row
CV08_10 <- rbind(CV08_Gini,CV10_Gini)

#declare panel data
CV08_10 <- plm.data(CV08_10, index=c("cname", "year"))

````

The question of whether ethnic diversity might have an effect on political rights was inspired by several interrelated pieces of research. First, of the vast body of academic literature attempting to define democracy and the various frameworks that have been constructed to do so, we found ourselves drawn to Wolfgang Merkel's conceptual framework of "embedded democracy". This framework characterizes stable constitutional by their internal and external "embeddedness". Internal embeddedness is defined by the simultaneous independence and interdependence of what Merkel calls the "five partial regimes" of democracy. These are: \itemize{
\item{a democratic electoral regime} 
\item{political rights of participation} 
\item{civil rights}
\item{horizontal accountability} 
\item{the effective power to govern by the democratically elected representatives} 
}

According to the theory, internal embeddedness refers to the qualities of these as simultaneously independent from each other (independence) and support each other (interdependence). 

External embeddedness, on the other hand, refers to the embeddedness of this webbed structure in spheres of enabling conditions that protect it from inner and outer destabilizers and shocks, such as a developed economy, fair distribution of resources, and prevention of extreme poverty, to name a few conditions. By his same definition, any democracy where the mutual embeddedness of the five partial regimes is no longer intact is labelled a "defective democracy" [-@Merkel2004].

Our second main source of inspiration was Paul Collier's studies on the effect of ethnic fractionalization, or the probability that any two individuals drawn from a given population will belong to different ethnic groups, on conflict, and how that effect differs between regime types. He shows that that, while higher ethnic fractionalization can actually play a stabilizing role in developed democracies, it can be quite harmful in authoritarian regimes, depending on the pattern of fractionalization that is present [-@Collier2000].

This got us to thinking: what sort of role might ethnic diversity, or fractionalization, play on each of Merkel's five partial regimes that qualify developed democracies? We decided to focus on its effect on political rights, which Merkel defines by two different qualifiers: freedom of speech and opinion, and the freedom of association, demonstration, and petition. He justifies their allotted role as a partial democratic regime by pointing out the important roles they play as an enabler to fair and organized democratic elections and in ensuring the complete development of political society, which in turn behaves as a check and balance on state institutions.

However, we did not think that this was all there was to the equation and so, based on the belief that the level of development of a country would likely play a significant role in explaining the impact of ethnic fractionalization on political rights and recent scholarly research suggesting that excessive inequalities undermine democratic political regimes [-@Acemoglu2006] and research showing that inequality perpetuated by ethnic divisions is particularly bad for economic growth relative to those perpetuated by other variables and  that ethnic fractionalization is an important determinant of institutional quality [-@CaseyOwen2010], we thought that ethnic fractionalization might play an important role in determining the level of Merkel's political rights when the country is at an advanced stage of development, which we define as the broad distribution of wealth in a country (low economic inequality). 

As a proxy for these factors, we adopted the following:\itemize{
\item{for political rights, Freedom House's freedom of the press indicator} 
\item{for ethnic diversity, Alesina et. al.'s ethnic fractionalization indicator} 
\item{and for inequality/level of development, the Gini coefficient}
}

We hypothesized that high ethnic fractionalization would have a negative impact on the preservation of political rights, as different groups compete to dominate the political discourse. We also thought that its impact might depend on the level of income inequality in a country: the less equal a country, the higher the negative impact high ethnic fractionalization will have on political rights. Conversely, the more equal a country, the higher the positive impact high ethnic fractionalization will have on political rights. Our main hypotheses were, then:

**H1: The more ethnically fractionalized a country is, the higher its negative impact on political rights if income inequality is high.**

**H2: When income inequality is low, the impact of high ethnic fractionalization on political rights is expected to be positive.**

We also included the variables of linguistic fractionalization, income (as measured by GDP per capita),  and judicial independence as controls.

All the variables used in this study are easily accessed in the Quality of Government Standard Dataset developed by the Quality of Government Institute from Gothenburg [@QualityofGovernment2015], the World Bank's World Governance Indicators dataset, and the World Development Indicators dataset, which is provided as an R package, also by the World Bank. For more information, please se our Github repository directly: https://github.com/imiller88/IsabelleandDiegosFinalResearchProject.

#Methodology
Due to the lack of availability of data spanning multiple years by our main independent variables, ethnic fractionalization and the Gini coefficient, we were only able to run simple linear models. Therefore, our primary model consists of pooled OLS regressions for the years 2008 and 2010, for which we were able to collect the most complete data. 

Our basic equation (from the pooled OLS model) was as follows:
\vspace{0.5cm}

*PoliticalRightsi= B0 + B1EthnicFraci + B2IncomeIneqi + B3EthnicFracxIncomeIneqi + B4GDPcapitai + B5Judiciaryindependencei + B6Voiceandaccounti +B7Linguisticfraci + Errori*

##*A side note on limitations*
After thoroughly cleaning the data of all unavailable data points, we were left with an n of 65 for 2008 and 67 for 2010. As seen in the images above, most African countries, smaller island nations, and all of South Asia have unfortunately been dropped from this sample due to lack of available Gini data; this means that, while there are some developing countries in Asia and Latin America included in the sample, it is disproportionately skewed towards Western and OECD countries. Therefore, we also include regression results run for the larger sample excluding the interacting variable (Gini coefficient).

Additionally, this study is not able to show any possible changes over time. In order to be able to draw firmer conclusions, it would be ideal to run a panel data model over the course of at least 20-100 years to allow for changes in ethnic fractionalization and inequality to take place due to migration, economic cycles, regime changes, and/or various other mitigating factors. 


#Data Exploration
In an attempt to identify any significant patterns, we first look at the distributions of our main variable of interest. Please note the skewness of available Gini Coefficients and the slight skewness of ethnic fractionalization.

````{r histograms, echo=FALSE, error=FALSE, message=FALSE}

pressfree10hist <- qplot(pressfreedom, data=CV10_Gini, geom="histogram") + geom_histogram(aes(fill = ..count..)) +
  scale_x_reverse() +
  labs(title="Press Freedom, 2010", x="Freedom (least to most free)", y="Countries in sample")
pressfree10hist

ethnicfrac10hist <- qplot(ethnicfrac, data=CV10_Gini, geom="histogram") + geom_histogram(aes(fill = ..count..)) +
  scale_x_reverse() +
  labs(title="Ethnic Fractionalization, 2010", x="Ethnic Fractionalization, heterogenous to homogenous",       y="Countries in sample")
ethnicfrac10hist

gini10hist <- qplot(Ginicoef, data=CV10_Gini, geom="histogram") + geom_histogram(aes(fill = ..count..)) + labs(title="Distribution of Available Gini Coefficients for 2010", x="Inequality (low to high)", y="Frequency")
gini10hist
````


Here we can also see that there seems to be a general negative correlation between both variables (the higher the fractionalization, the lower the press freedom, in accordance with our theory) but there is also a clear dip around an ethnic fractionalization value of 0.30 (a 0.30 chance that any two individuals selected from a given society will belong to different ethnic groups). 

\vspace{1cm}

````{r ggplots, echo=FALSE, error=FALSE, message=FALSE}
ggplot2::ggplot(CV10_Gini, aes(ethnicfrac, pressfreedom)) + geom_point() + geom_smooth() + theme_bw() + 
  ggtitle("Ethnic Fractionalization & Press Freedom 2010") +
  xlab("Ethnic fractionalization") +
  ylab("Press Freedom") + scale_y_reverse()

````
#OLS regression

When it comes to the inferential part of our analysis, its core is constituted by two OLS regressions with the chosen model specification for the years 2008 and 2010, for which we are able to collect the most complete data, as we have already said before. However, we also run OLS regressions for those two years without including the Gini coefficient in our model (due to the fact that when we include the Gini coefficient in our model, many countries drop from our sample since there are not data available for them), and using organizational freedom as (alternative) dependent variable. Below you can find separate tables for the estimate coefficients obtained from our different model specifications for each of the two years that we are analyzing. We also run a first difference model for the years 2008 and 2010, and 1-year OLS models with the quadratic term as well as the log of ethnic fractionalization. Nevertheless, due to the insignificance of the estimate coefficients obtained, and for the sake of clearness, we decided not to include them in this final paper. 

```{r, echo=FALSE, warning=FALSE, error=FALSE, message=FALSE, tidy=TRUE, results="asis"}

#setting data as panel data
CV08_10 <- plm.data(CV08_10, index=c("cname", "year"))

OLS08 <- plm(pressfreedom ~ ethnicfrac * Ginicoef + GDPpercapita + langfrac + judindep, data = CV10_Gini, model="pooling", index=c("cname"))

OLS10 <- plm(pressfreedom ~ ethnicfrac * Ginicoef + GDPpercapita + langfrac + judindep, data = CV10_Gini, model="pooling", index=c("cname"))

labels_IVs <- c("Ethnic Fractionalization", "GDP per capita", "Linguistic Fractionalization", "Judicial Independence", ("Intercept"))

label_DV <- c("Freedom of the press")

stargazer::stargazer(OLS08, OLS10, dep.var.labels=label_DV, covariate.labels = labels_IVs, title = "Regression Results: Year 2008 (1) and Year 2010 (2)", digits = 2, type = "html", header = FALSE)


```



#Conclusion and Grounds for Further Analysis
As previously mentioned, in order to truly be able to draw conclusive relationships between the variables at hand, a larger dataset spanning a much greater time period would be needed. Unfortunately, data on ethnic fractionalization going back large periods of time is difficult to come by, and we were not able to include it in this study.

\newpage

##Bibliography