---
title: "Third Pair Assignment"
author: "Diego Fernández Fernández and Isabelle Miller"
date: "19 Apr 2016"
output: pdf_document
bibliography: Citations2.bib
---

#Research Question
Our Collaborative Research Paper aims at shedding some light on what the effects of ethnic fractionalization on political rights are. In this sense, we have two main hypothesis: 

**H1: The more ethnically fractionalized a country is, the higher its negative impact on political rights if income inequality is high.**

**H2: When income inequality is low, the impact of high ethnic fractionalization on political rights is expected to be positive.**

For more information regarding the theoretical underpinnings of our research proposal, you can have a look at [https://github.com/imiller88/IsabelleandDiegosFinalResearchProject/blob/master/Assignment2/Assignment2_ResearchProposal.Rmd]

```{r}
summary(cars)
```

You can also embed plots, for example:

```{r, echo=FALSE}
plot(cars)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
